A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-25 paper.tex
Sections 2 & 3 are hopefully complete & consistent now

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$\langle s, r \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$\langle s, r \rangle, T$} \Comment{Look for a candidate}
        \IIf{$r = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$\langle s, r \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$\langle s, r \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$\langle s, r \rangle, T$}
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land$ \Call{treeFormula}{$T$} }
            \IIf{$sol = \texttt{unsat}$} 
                \Return $\emptyset$ \Comment{No candidate exists}
            \EndIIf
            \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$\langle s, r\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle r', s'\rangle \gets $ \Call{outcome}{$\langle s, r\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{\textoverline{solveAbstract}}{$\langle r', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Require $\sigma(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $\sigma(X_T) \land B^M \equiv \bot$
        \Function{learn}{$\sigma, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with max rank
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$\sigma(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$\sigma, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets E$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
                \State \Call{solveAbstract}{$\langle I, k \rangle, \emptyset$}
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Function{solveAbstract}{$\langle s, r \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$\langle s, r \rangle, T$} \Comment{Look for a candidate}
        \IIf{$r = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$\langle s, r \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$\langle s, r \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{u2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u2}
        \Function{findCandidate}{$\langle s, r \rangle, T$}
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land$ \Call{treeFormula}{$T$} }
            \If{$sol = \texttt{unsat}$} 
                \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                \State \Call{learn}{$\sigma, T$}
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{u3}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u3}
        \Function{verify}{$\langle s, r\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle r', s'\rangle \gets $ \Call{outcome}{$\langle s, r\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{\textoverline{solveAbstract}}{$\langle r', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

The unbounded synthesis algorithm (Algorimthm \ref{alg:unbounded}) effectively
consists of two communicating solvers: the counterexample-guided bounded
reachability solver and the unbounded solver based on incremental induction.
Fortunately, correctness of both solvers can be established largely
independently, which simplifies the proof of correctness of the overall
algorithm.

@@

\subsection{Proof of \textsc{learn}}

We prove that postconditions of \textsc{learn} are satisfied assuming that its
preconditions hold.

Line~(4) splits the tree $T$ into $T_1$ and $T_2$, such that $T_2$ has
depth 1.  Consider formulas $f_1=\sigma(X_T) \land \textsc{treeFormula}(T_1)$
and $F_2 = \textsc{treeFormula}(T_2)$.  These formulas only share variables
$X_n$.  Their conjunction $F_1 \land F_2$ is unsatisfiable, as by construction
any solution of $F_1 \land F_2$ also satisfies $\sigma(X_T) \land
\textsc{treeFormula}(T)$, which is unsatisfiable (precondition (b)).  Hence the
interpolation operation is defined for $F_1$ and $F_2$.

Intuitively, the interpolant computed in line~(5) overapproximates the set of 
states reachable from $\sigma$ by following the tree from the root node to $n$,
and underapproximates the set of states from which the controller loses against 
tree $T_2$.

Formally, $\II$ has the property $\II \land F_2 \equiv \bot$.  Since $T_2$ is of depth
1, this means that the controller cannot force the
game into $W^m[|rank|(n-1)]$ when playing against counterexample moves in $T_2$.  Hence, 
$\II \cap Cpre(W^m[|rank|(n)]) = \emptyset$.  Furthermore, since the
may-invariant holds (assumption (c)), $\II \cap Cpre(i) = \emptyset$, for
all $i<|rank|(n)$.  Hence, removing $\II$ from all $W^m[i], i\leq |rank|(n)$
in line~(6) preserves the may-invariant, thus satisfying the first post-condition.

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any assignment to 
$X_n$ that satisfies $\sigma(X_T) \land |treeFormula|(T_1)$ also satisfies $\II$.  Hence, removing $\II$ 
from $W_m[|rank|(n)]$ makes $\sigma(X_T) \land |treeFormula|(T_1)$ unsatisfiable, 
and hence all preconditions of the recursive invocation of \textsf{learn} in line~(7) 
are satisfied.

At the second las recursive call to \textsf{learn}, tree $T_1$ is empty, $n$ is the root node,
$|treeFormula|(T_1) \equiv W^m[|rank|(T)](X^T)$; hence 
$\sigma(X_T) \land |treeFormula|(T_1) \equiv \sigma(X_T) \land W^m[|rank|(T)](X^T) \equiv \bot$.
Thus the second postcondition of \textsf{learn} holds.

@@

We must prove that $|checkRank|$ terminates and that upon 
termination its postcondition holds, i.e., state $s$ is added to $W^M$ if 
its rank is no greater than $k$ or is removed from $W^m[k]$ otherwise.  
Termination follows from completeness of couterexample-guided search, which 
terminates after enumerating all possible opponent moves in the worst case.

Assume that $|rank|(s) > k$.  This means that at some point the algorithm
discovers a counterexample tree of rank $k$ that is losing for the controller 
from state $s$.  The algorithm then invokes the $|learn|$ method, which 
eliminates $s$ from $W^m[k]$.  Alternatively, if $|rank|(s) \leq k$, a 
counterexample tree losing for the environment will be discovered and 
subsequent call to $\overline{|learn|}$ adds $s$ to $W^M$.

>>>>

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$\langle s, k \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$\langle s, k \rangle, T$} \Comment{Look for a candidate}
        \IIf{$k = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$\langle s, k \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$\langle s, k \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$\langle s, k \rangle, T$}
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land$ \Call{treeFormula}{$T$} }
            \IIf{$sol = \texttt{unsat}$} 
                \Return $\emptyset$ \Comment{No candidate exists}
            \EndIIf
            \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$\langle s, k\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$\langle s, k\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{\textoverline{solveAbstract}}{$\langle k', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Require $\sigma(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $\sigma(X_T) \land B^M$
        \Function{learn}{$\sigma, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with max rank
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$\sigma(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$\sigma, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets E$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $\lnot s \in B^m[k]$ if there is a winning controller strategy of length $k$ starting at $s$
        \Ensure $s \in B^M$ if there is a winning environment strategy of length $k$ starting at $s$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkRank}{$s, k$}
            \State \Return \Call{solveAbstract}{$\langle s, k \rangle, \emptyset$}
        \EndFunction
        \algstore{u2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u2}
        \Function{solveAbstract}{$\langle s, k \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$\langle s, k \rangle, T$} \Comment{Look for a candidate}
        \IIf{$k = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$\langle s, k \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$\langle s, k \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{u3}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u3}
        \Function{findCandidate}{$\langle s, k \rangle, T$}
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land$ \Call{treeFormula}{$T$} }
            \If{$sol = \texttt{unsat}$} 
                \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                \State \Call{learn}{$\sigma, T$}
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{u4}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u4}
        \Function{verify}{$\langle s, k\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$\langle s, k\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{\textoverline{solveAbstract}}{$\langle k', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

The unbounded synthesis algorithm (Algorithm \ref{alg:unbounded}) effectively
consists of two communicating solvers: the counterexample-guided bounded
reachability solver and the unbounded solver based on incremental induction.
Fortunately, correctness of both solvers can be established largely
independently, which simplifies the proof of correctness of the overall
algorithm.

@@

\subsection{Proof of \textsc{\textoverline{learn}}}

We prove that postconditions of \textsc{\textoverline{learn}} are satisfied
assuming that its preconditions hold.

Line~(11) splits the tree $T$ into $T_1$ and $T_2$, such that $T_2$ has depth
1.  Consider formulas $f_1=\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ and $F_2 =
\textsc{\textoverline{treeFormula}}(T_2)$.  These formulas only share variables
$X_n$.  Their conjunction $F_1 \land F_2$ is unsatisfiable, as by construction
any solution of $F_1 \land F_2$ also satisfies $\sigma(X_T) \land
\textsc{treeFormula}(T)$, which is unsatisfiable (precondition (b)).  Hence the
interpolation operation is defined for $F_1$ and $F_2$.

Intuitively, the interpolant computed in line~(13) overapproximates the set of
states reachable from $\sigma$ by following the tree from the root node to $n$,
and underapproximates the set of states from which the environment loses
against tree $T_2$.

Formally, $\II$ has the property $\II \land F_2 \equiv \bot$.  Since $T_2$ is
of depth 1, this means that the environment cannot force the game into
$B^m[\textsc{rank}(n)]$ playing against the counterexample moves in $T_2$.
Hence, $\II \cap Upre(B^m[\textsc{rank}(n)]) = \emptyset$.  Furthermore, since
the may-invariant holds, $\II \cap Upre(i) = \emptyset$, for all $i <
\textsc{rank}(n)$.  Hence, removing $\II$ from all $B^m[i], i\leq
\textsc{rank}(n)$ in line~(15) preserves the may-invariant, thus satisfying the
first post-condition.

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any
assignment to $X_n$ that satisfies $\sigma(X_T) \land |treeFormula|(T_1)$ also
satisfies $\II$.  Hence, removing $\II$ from $B^m[\textsc{rank}(n)]$ makes
$\sigma(X_T) \land |treeFormula|(T_1)$ unsatisfiable, and hence all
preconditions of the recursive invocation of \textsc{\textoverline{learn}} in
line~(17) are satisfied.

At the second last recursive call to \textsc{\textoverline{learn}}, tree $T_1$
is empty, $n$ is the root node, $\textsc{\textoverline{treeFormula}}(T_1)
\equiv B^m[\textsc{rank}(T)](X^T)$; hence $\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1) \equiv \sigma(X_T) \land
B^m[\textsc{rank}(T)](X^T) \equiv \bot$.  Thus the second postcondition of
\textsc{\textoverline{learn}} holds.

@@

We must prove that \textsc{checkRank} terminates and that upon termination its
postcondition holds.

We must prove that $\textsc{checkRank}$ terminates and that upon termination
its postcondition holds, i.e., state $s$ is removed from $B^m[k]$ if there is a
winning controller strategy on the bounded safety game of rank $k$ or it is
added to $B^M$ otherwise. Termination follows from completeness of
counterexample-guided search, which terminates after enumerating all possible
opponent moves in the worst case.

Assume that there is a winning strategy for the controller at rank $k$. This
means that at some point the algorithm discovers a counterexample tree of rank
$k$ for which the environment cannot force into $E$. The algorithm then invokes
the \textsc{\textoverline{learn}} method, adds $s$ to $B^M$.  Alternatively, if
there is a winning strategy for the environment at rank $k$ then a
counterexample losing for the controller will be found. Subsequently
\textsc{learn} will be called and $s$ eliminated from $B^m[k]$.
